---
title: "Phase3_ADL_Modeling"
author: "Sam Sithimolada"
date: "`r Sys.Date()`"
output: html_document
---

# AAA Washington Case Overview

Following successful regression and ARIMA modeling phases, AAA Washington retained our consulting services to develop a more comprehensive forecasting model. Our goal in this phase is to build a time series regression model that accounts for both seasonal and economic influences on emergency road service call volume.

We incorporated insights from both the operations and dispatch teams. First, engineers believe that most vehicles are optimized to operate at around 60°F. Second, the emergency services team observed sharper increases in call volume when temperatures dip slightly below the 30s, suggesting a nonlinear relationship. Additionally, Michael DeCoria believes emergency service demand is partially driven by broader economic conditions, and has recommended the Washington State unemployment rate as a relevant cyclical indicator.

## 1. Exploratory Data Analysis

We began by importing the dataset `aaa_washington.csv` and creating time series objects for call volume, average monthly temperature, and unemployment rate. Plots of each series show a strong seasonal pattern, especially in temperature and call volume. The unemployment rate exhibits cyclical fluctuations and possible seasonality, with a trend resembling a quadratic form.

```{r}
data = read.csv("aaa_washington.csv")
call.ts = ts(data$Calls, start = c(1988, 8), frequency = 12)
temp.ts = ts(data$Temp, start = c(1988, 8), frequency = 12)
rate.ts = ts(data$Rate, start = c(1988, 8), frequency = 12)
```

```{r Call Vol Graph}
library(forecast)
autoplot(call.ts, main = "Monthly AAA Call Volume", ylab = "Call Volume")
```

```{r Temp Graph}
autoplot(temp.ts, main = "Monthly Washington State Avg Temp", 
         ylab = "Avg Temp (F)")
```

```{r Unemployment Rate Graph}
autoplot(rate.ts, main = "Monthly Washington State Unemployment Rate", 
         ylab = "Unemployment Rate")
```

Visual inspection confirms that call volume dips mid-year and rises again toward winter, while temperatures follow an opposite seasonal trajectory. The unemployment rate shows broader cyclical movement.

## 1.1. Temperature Transformation

To reflect engineering insights, we created a new temperature variable by repositioning the scale around 60°F and taking the absolute deviation from this optimal operating point. This transformation revealed a more linear relationship between temperature deviation and call volume compared to the original temperature series, which displayed a nonlinear, possibly quadratic relationship.

```{r}
newtemp.ts = abs(temp.ts - 60)

#plot(newtemp.ts, call.ts) #gets messy because it plots years

#use as.vector(ts) to plot
#temp vs calls
plot(as.vector(temp.ts,), as.vector(call.ts),
     main = "Call Vol v Temp", xlab = "Temp", ylab = "call volume")

#newtemp vs calls
plot(as.vector(newtemp.ts,), as.vector(call.ts),
     main = "Call Vol v Temp Deviated from 60F", xlab = "Temp from 60", ylab = "call volume")
```


## 1.2. Lag Analysis: Economic Indicator

Cross-correlation analysis between the unemployment rate and call volume showed that the unemployment rate leads call volume, with the strongest correlation at a lag of 11 months. A secondary, smaller peak appeared at lag 3. Based on this, we created lagged unemployment variables for further modeling.

```{r, cross-correlation between unemployment rate and calls}

#Ccf(X, y, lags)
Ccf(rate.ts, call.ts) 

#use +-12 lags because they are insignificant beyond that (no spikes)
ccf.val = Ccf(rate.ts, call.ts, 12)

#filter only negative lags
negative_lags = ccf.val$lag<0
negative_acf = ccf.val$acf[negative_lags]
negative_lags_only = ccf.val$lag[negative_lags]

#finds index of max ccf in negative lags (t/f)
index = which(negative_acf == max(negative_acf))

#retrieves lag corresponding to max correl for which index is true
negative_lags_only[index]
```

# 2. Modeling: Univariate Lagged Models

We developed two simple regression models using lagged unemployment rates. Both models show the lagged unemployment variables to be statistically significant at 5%. However, their explanatory power was limited:
- Adjusted R²: 7.35%
- RMSE: ~1546
- MAPE: 6.16%

These models alone are insufficient for forecasting call volume with high accuracy.

```{r, univariate lag models}
#lag unemployment rate variables
leadrate3 = lag(rate.ts, -3)
leadrate11 = lag(rate.ts, -11)

#tslm models with lag3 and lag10 of unemployment rate on call volume
model_lag3 = tslm(call.ts~leadrate3)
model_lag11 = tslm(call.ts~leadrate11)
```

```{r, Checking Lag Unemployment Rate Variable Significance}
#summary table
summary(model_lag3)
summary(model_lag11)

#rmse
accuracy(model_lag3)[2]
accuracy(model_lag11)[2]

#mape
accuracy(model_lag3)[5]
accuracy(model_lag10)[5]
```

## 2.1 Multivariate Time Series Regression Model

Next, we combined the transformed temperature variable and the 11-month lagged unemployment rate into a multiple regression model.

```{r, Multiple Regression Model with NewTemp and Lagged Unemployment Rate Variables}
model_temp_lag11 = tslm(call.ts ~ newtemp.ts + leadrate11)

summary(model_temp_lag11)

#rmse
accuracy(model_temp_lag11)[2]

#mape
accuracy(model_temp_lag11)[5]
```

In this model:
- The temperature deviation was statistically significant (p < 0.05).
- The unemployment rate (leadrate11) was not significant individually, but improved overall model performance.

Model performance improved substantially:
- Adjusted R²: 57.11%
- RMSE: 1041.91
- MAPE: 3.58%

## 3. Residual Diagnostics

Diagnostic checks on the residuals indicate that the model may not fully satisfy all assumptions:
- Residuals appear normally distributed.
- ACF plot shows significant autocorrelation at lag 1 and 2, suggesting residual dependence.
- Durbin-Watson test confirms the presence of serial correlation (p < 0.05).
- Durbin-Watson Test:
    H_0: Errors have no autocorrelation
    H-1: Errors have autocorrelation

```{r}
checkresiduals(model_temp_lag11)
dwtest(model_temp_lag11)
```

## 4. Conclusion and Final Recommendation

The most effective model combines the transformed temperature deviation from 60°F and the unemployment rate lagged by 11 months. While the unemployment variable is not statistically significant in isolation, its inclusion enhances the model’s predictive power, resulting in a much lower forecast error. With a MAPE of just 3.58% and an adjusted R² exceeding 57%, this model offers a practical and reliable forecasting tool for emergency road service demand.

However, we recommend further refining the model to address the residual autocorrelation. Incorporating ARIMA-style residual corrections or switching to an ADL-ARIMA hybrid may further improve model robustness and accuracy in future work.
